Topological Equivariant Artist Model

### 2025-04-03 paper/main.tex
fixed xi figure
talk version
spivak table

\usepackage[outputdir=.]{minted}
\setminted[python]{fontsize=\scriptsize,
                   linenos,
                   numbersep=8pt,
                   frame=lines,
                   autogobble,
                   framesep=3mm,
                   breaklines=True}
% *** ALIGNMENT PACKAGES ***
\usepackage{multicol}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabulary}
% IEEEtran contains the IEEEeqnarray family of commands

@@

\title{Topological Equivariant Artist Model for Visualization Library Architecture}
% author names and IEEE memberships
\author{Hannah~Aizenman, Mikael Vejdemo-Johansson, Thomas~Caswell, and~Michael~Grossberg,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem H. Aizenman is with the department of Computer Science, The Gradaute Center, CUNY.
\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: haizenman@gradcenter.cuny.edu,
\IEEEcompsocthanksitem M. Grossberg is with the department of Computer Science, City College of New York, CUNY.
E-mail: mgrossberg@ccny.cuny.edu
\IEEEcompsocthanksitem Mikael Vejdemo-Johansson is with the department of Mathematics, CUNY College of Staten Island.
\protect\\
E-mail: mvj@math.csi.cuny.edu
\IEEEcompsocthanksitem Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab
\protect \\
E-mail: tcaswell@bnl.gov}% <-this % stops an unwanted space
\thanks{Manuscript received X XX, XXXX; revised X XX, XXXX.}
}

@@

The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure data types as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

@@

\subsubsection{\textcolor{fiber}{Data Field Types: Fiber Space \dfiber}}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, visualization researchers traditionally describe equivariance as the preservation of field structure, which is based on the field type. Spivak shows that data typing can be expressed in a categorical framework in his fiber bundle formulation of tables in relational databases \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. In this work, we adopt Spivak's definitions of \textit{type specification}, \textit{schema}, and \textit{record} because that allows us to use a dimension agnostic named typing system for the fields of our dataset that is consistent with the abstraction we are using to express the continuity. Spivak introduces a \textit{type specification} as a bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is a set of data types $T \in \textbf{DT}$ and the total space $\mathscr{U}$ is the disjoint union of the domains of each type \[\mathscr{U} = \bigsqcup_{T \in \textbf{DT}} \pi^{-1}(T)\] such that each element $x$ in the domain $\pi^{-1}(T)$ is one possible value of an object of type $T$ \cite{spivakSimplicialDatabases2009}. For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers and $x=3 \in \mathbb{Z}$ is the value of one $\texttt{int}$ object.

Since many fields can have the same datatype, Spivak formally defines a mapping from field name to field data type, akin to a database schema \cite{ullmanFirstCourseDatabase2008}. According to Spivak, a \textit{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSimplicialDatabases2009}. The function $\sigma$ is composed with $\pi$ such that $\pi^{-1}(\sigma(C)) \subseteq \mathscr{U}$; this composition induces a domain bundle $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ that associates a field name $c \in C$ with its corresponding domain $\pi^{-1}_{\sigma}(C) \subseteq \mathscr{U}_{\sigma}$.
\begin{definition} A \textbf{record} is a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$.
\end{definition}
Spivak then describes tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from an indexing space $\dbase$ to the set of all possible records $\Gamma^{\pi}(\sigma)$ on the schema bundle, and his notion of a table generalizes to our notion of a data container.

To build on the rich typing system provided by Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the space of all possible data records
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$.
Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accessed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a mathematical space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

>>>>

\usepackage[]{minted}
\setminted[python]{fontsize=\scriptsize,
                   linenos,
                   numbersep=8pt,
                   frame=lines,
                   autogobble,
                   framesep=3mm,
                   breaklines=True}
% *** ALIGNMENT PACKAGES ***
\usepackage{multicol}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabulary}
\usepackage{tabularray}
% IEEEtran contains the IEEEeqnarray family of commands

@@

\title{Topological Equivariant Artist Model}
% author names and IEEE memberships
\author{Hannah~Aizenman, Mikael Vejdemo-Johansson, Thomas~Caswell, and~Michael~Grossberg,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem H. Aizenman is with the department of Computer Science, The Gradaute Center, CUNY.
\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: haizenman@gradcenter.cuny.edu,
\IEEEcompsocthanksitem M. Grossberg is with the department of Computer Science, City College of New York, CUNY.
E-mail: mgrossberg@ccny.cuny.edu
\IEEEcompsocthanksitem Mikael Vejdemo-Johansson is with the department of Mathematics, CUNY College of Staten Island.
\protect\\
E-mail: mvj@math.csi.cuny.edu
\IEEEcompsocthanksitem Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab
\protect \\
E-mail: tcaswell@bnl.gov}% <-this % stops an unwanted space
\thanks{Manuscript received X XX, XXXX; revised X XX, XXXX.}
}

@@

The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and morphisms $\opensetc_i \xrightarrow{\hat{\phi}} \opensetc_j$\cite{riehlCategoryTheoryContext}.

\note{move $\hat{\phi}$ here}

The composability property expresses that the morphism is transitive, while associativity expresses that the morphisms can be curried in various equivalent groupings. By formally specifying the properties of the topological structure data types as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

@@

\subsubsection{\textcolor{fiber}{Field Data Types: Fiber Space \dfiber}}
\label{sec:atct:fb:fiber}
As mentioned in \autoref{sec:related-work:equivariance}, visualization researchers traditionally describe equivariance as the preservation of field structure, which is based on the field type. Spivak shows that data typing can be expressed in a categorical framework in his fiber bundle formulation of tables in relational databases \cite{spivakDatabasesAreCategories2010,spivakSimplicialDatabases2009}. We adopt Spivak's definitions of \textit{type specification}, \textit{schema}, \textit{record}, and \textit{table} because that allows us to use a dimension agnostic named typing system for the dataset fields that is consistent with the abstraction we are using to express the continuity.

Spivak introduces a \textit{type specification} as a bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is a set of data types $T \in \textbf{DT}$ and the total space $\mathscr{U}$ is the disjoint union of the domains of each type $\mathscr{U} = \bigsqcup_{T \in \textbf{DT}} \pi^{-1}(T)$ such that each element $x$ in the domain $\pi^{-1}(T)$ is one possible value of an object of type $T$ \cite{spivakSimplicialDatabases2009}. For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers and $x=3 \in \mathbb{Z}$ is the value of one $\texttt{int}$ object. Spivak's analogue of a database \textit{schema}\cite{ullmanFirstCourseDatabase2008} is the \textit{simple schema of type $\pi$} which is defined as consisting of the pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSimplicialDatabases2009}. The function $\sigma$ is composed with $\pi$ such that $\pi^{-1}(\sigma(C)) \subseteq \mathscr{U}$; this composition induces a domain bundle $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ that associates a field name $c \in C$ with its corresponding domain $\pi^{-1}_{\sigma}(C) \subseteq \mathscr{U}_{\sigma}$. He defines a \textit{record} as a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ from column name to type domain and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$. Spivak then describes tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from an indexing space $\dbase$ to the set of all possible records $\Gamma^{\pi}(\sigma)$ on the schema bundle, and his notion of a table generalizes to our notion of a data container.

In summary, Spivak defines types as maps between type name (e.g. $temperature \in C$), data type ($\texttt{float} \in \texttt{DT}$), and type domain $\mathbb{R}\subseteq \mathscr{U_{\sigma}} \subseteq \mathscr{U}$). Records are maps from name to value ($\textit{temperature}:152.0^{\circ} F$) and tables are maps from index to record \cite{spivakSimplicialDatabases2009}:
\begin{table}[H]
  \begin{equation*}
    \begin{tabular}{|c|c|c|}
      \hline
       types and schema & \textcolor{fiber}{record} & \textcolor{section}{table}\\
      \hline
      \begin{tikzcd}[ampersand replacement=\&, sep=huge]
        \textcolor{total}{\mathcal{U}_\sigma} \arrow[d, "\pi_{\sigma}"', color=total, two heads] \arrow[r, dotted] \arrow[rd, dotted] \& \textcolor{total}{\mathcal{U}} \arrow[d, "\pi"', color=total, two heads] \\
        \textcolor{base}{C} \arrow[r, "\sigma"']                                                                                   \& \textcolor{base}{\texttt{DT}}
      \end{tikzcd}
    &
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
      \textcolor{total}{\mathcal{U}_{\sigma}} \arrow[d, "\pi_{\sigma}"', color=total, two heads] \\
      \textcolor{base}{C} \arrow[u, "\delementc"', dashed, bend right, color=fiber]
      \end{tikzcd}
    &
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
      \dfiberc\coloneqq \textcolor{total}{\Gamma^{\pi}(\sigma)} \arrow[d, "\pi"', color=total, two heads] \\
      \dbasec \arrow[u, "\dsectionc"', dashed, bend right, color=section]
      \end{tikzcd}\\
      \hline
    \end{tabular}
    \end{equation*}
    \caption{}
    \label{tab:fiber_specification}
  \end{table}

Building on Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the space of all possible data records:
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accessed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a mathematical space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.
